---
title: "clean-download-EDA"
author: "Anna Spiers"
date: "7/9/2020"
output: html_document
---

1) Download NEON's carabid pitfall trap data product  
2) EDA  
3) Clean data and save locally

```{r, message=F}
library(ggplot2)
library(dplyr)
library(stringr) #word()
library(lubridate)
library(neonUtilities) 

# Check that neonUtilities package is up-to-date
if (sessionInfo()$otherPkgs$neonUtilities$Version != "1.3.6") {
    library(devtools) #install_github()
    install_github("NEONScience/NEON-utilities/neonUtilities")
    # restart R
}

```

#### Download

At first, the `neonUtilities::loadByProduct` function downloaded only a partial dataset (20 MB). Then, I updated `neonUtilities` via CRAN and the error described in this [bug report](https://github.com/NEONScience/NEON-utilities/issues/94) occurred. The resolution to that bug is to download the latest `neonUtilities` version from github. Then restart R. Now, all 67.4 MB of data is downloaded using the `neonUtilities::loadByProduct` function. 
```{r, message=F}

# # Download carabid pitfall trap survey data 
if (!file.exists("data/carabid_raw.rds")) {
    carabid_raw <- loadByProduct(dpID = "DP1.10022.001", check.size = T) #67.4 MB
    saveRDS(carabid_raw, "data/carabid_raw.rds")
}
carabid_raw <- readRDS("data/carabid_raw.rds")
list2env(carabid_raw, .GlobalEnv)
rm(carabid_raw)
```


#### EDA

Look at the data composition
```{r}
bet_parataxonomistID %>% count(taxonRank, sort=TRUE) #look into morphospecies too
bet_parataxonomistID %>% count(siteID, sort=TRUE) # 47 sites
bet_sorting %>% filter(sampleType == "carabid" | sampleType == "other carabid") %>%
    count(taxonRank, sort=TRUE) #look into morphospecies too
```
If all individual beetles collected are in `bet_sorting`, then how can there be more species in in `bet_parataxonomistID`? Is there more refined identification in the pinning process when `bet_parataxonomist` is created?

Proportion of traps that are missing out of total possible. sampleCollected=N means sample/trap was not collected on a given collection bout. 
```{r}
bet_fielddata %>%
    ggplot() +
    geom_bar(aes(x=siteID, fill=sampleCollected), position = "fill") +
    theme(axis.text.x = element_text(angle = 90))
```
Every site is missing traps; some are missing a substantial number. This supports our choice to standardize abundance by trap-night.


Years of data available by site
```{r}
bet_fielddata %>%
    mutate(year = year(setDate)) %>%
    group_by(siteID, year) %>%
    summarize(first_set = range(setDate, na.rm=T)[1],
              last_collect = range(setDate, na.rm=T)[2]) %>%
    ggplot() +
    geom_linerange(aes(x=siteID, ymin=first_set, ymax=last_collect)) +
    coord_flip()
#x-axis is year, y-axis is site
```


Count at each site across collection bouts
```{r}
bet_sorting %>%
    filter(sampleType == "carabid" | sampleType == "other carabid") %>%
    select(domainID, siteID, plotID, trapID, setDate, collectDate, 
           individualCount, sampleID, subsampleID) %>%
    mutate(plot_trap = paste0(plotID, "_", trapID, sep=""),
           trapnight = as.numeric(collectDate - setDate, units="days"),
           col_DOY = yday(collectDate),
           year = as.character(year(collectDate))) %>%
    group_by(siteID, year, col_DOY) %>%
    summarize(n=sum(individualCount)) %>%
    ggplot() + 
        geom_line(aes(x=col_DOY, y=n, col=year)) +
    facet_wrap(~siteID)
```
`bet_parataxonomist` and `bet_sorting` have NA's in `trapID` variable. This may be important in understanding how many traps were collected in each bout at a site and how many nights the traps were left out for. However, `bet_fielddata` also has trap-level entries - check this out. Why are there blank `sampleID` values when a `subsampleID` value exists.

Count at each site across years
```{r}
bet_sorting %>%
    filter(sampleType == "carabid" | sampleType == "other carabid") %>%
    select(domainID, siteID, plotID, trapID, setDate, collectDate, 
           individualCount, sampleID, subsampleID) %>%
    mutate(plot_trap = paste0(plotID, "_", trapID, sep=""),
           trapnight = as.numeric(collectDate - setDate, units="days"),
           year = year(collectDate)) %>%
    group_by(siteID, year) %>%
    summarize(n=sum(individualCount)) %>%
    ggplot(aes(x=year, y=n)) + 
        geom_line() +
    facet_wrap(~siteID)
```

```{r}
bet_sorting %>%
    filter(sampleType == "carabid") %>%
    select(domainID, siteID, plotID, trapID, setDate, collectDate,
                         sampleID, subsampleID, remarks) %>%
    left_join(bet_fielddata %>%
                  select(domainID, siteID, plotID, trapID, setDate, collectDate, sampleCollected, sampleID)) %>%
    filter(is.na(trapID)) #1185 rows have trapID == NA
    
# Join dataframes using bet_fielddata's trapID column, since bet_sorting has NAs in the trapID column
bet_sorting %>%
    filter(sampleType == "carabid"| sampleType == "other carabid") %>%
    select(domainID, siteID, plotID, setDate, collectDate,
                         sampleID, subsampleID, remarks) %>%
    left_join(bet_fielddata %>%
                  select(domainID, siteID, plotID, trapID, setDate, collectDate, sampleCollected, sampleID)) %>%
    filter(is.na(trapID)) #0 rows
# That seems to work! Now we know to use trap info from bet_sorting



```

The NEON beetle user guide says a "record from from bet_sorƟng may have zero (if no contents of the
subsampleID pinned) or mulƟple child records in bet_parataxonomistID depending on the number of individuals selected for pinning from each subsampleID".  
Can we find subsamples where the scientific names differ between bet_sorting and bet_parataxonomist?
```{r}
# First, look at a singular subsample
# One row per subsample in bet_sorting
bet_parataxonomistID %>% count(subsampleID, sort=T) %>% head(5)
# Take most abundant subsampleID
bet_sorting %>% filter(subsampleID=="xZ4ZeMJQZ4IAvxkYw5JSL65p8nHS7rDz3HkqrgCiwuY=") %>%
    select(scientificName)
bet_parataxonomistID %>% filter(subsampleID=="xZ4ZeMJQZ4IAvxkYw5JSL65p8nHS7rDz3HkqrgCiwuY=") %>%
    distinct(scientificName)
bet_sorting %>% count(subsampleID) %>% filter(n>1)

# Now, identify rows in bet_parataxonomistID without a match in bet_sorting
anti_join(bet_parataxonomistID %>% select(subsampleID, scientificName),
          bet_sorting %>% filter(sampleType == "carabid"| sampleType == "other carabid") %>%
              select(subsampleID, scientificName))
# 3854 rows in bet_parataxonomistID do not match a row in bet_sorting. That's a substantial size. What should we do with this information? That means that from sorting to pinning, parataxonomist species ID's change for 3854 individual beetles

```

```{r}
bet_sorting %>% 
  select(collectDate, siteID, domainID, scientificName, individualCount) %>%
  mutate(month = format(collectDate, "%Y-%m")) %>% 
  mutate(month = as.Date(paste(month, "01", sep="-"))) %>%
  group_by(scientificName, month, siteID, domainID) %>%
  summarize(count = sum(individualCount, na.rm = TRUE)) %>%
  ungroup()

bet_sorting %>% 
  select(collectDate, siteID, plotID, domainID, trapID, scientificName, individualCount )%>%
  mutate(month = format(collectDate, "%Y-%m")) %>% 
  mutate(month = as.Date(paste(month, "01", sep="-"))) %>%
  group_by(scientificName, month, siteID, domainID)
```

According to the NEON beetle user guide, specimen not identified to species are identified to morphospecies. Is this reflected in the data?
```{r}
bet_sorting %>%
    filter(sampleType == "carabid"| sampleType == "other carabid") %>%
    count(taxonRank, sort=T) 

# What is a "speciesgroup" taxonRank?
bet_sorting %>%
    filter(sampleType == "carabid"| sampleType == "other carabid") %>%
    filter(taxonRank == "speciesgroup") %>%
    select(siteID, taxonID, scientificName, morphospeciesID) #same scientificname for all, spans 3 sites

bet_sorting %>%
    filter(sampleType == "carabid"| sampleType == "other carabid",
           taxonRank != "species" & taxonRank != "subspecies",
           is.na(morphospeciesID)) %>%
    select(taxonRank, scientificName, morphospeciesID) 
# 2783 records where parataxonomist did not identify to species, but no morphospeciesID is offered. 
# What do we do with this information?
```

#### Clean

Create a dataframe where every row is an individual beetle that has been collected by NEON.
```{r, message=F}
carabid_clean <-  bet_sorting %>%
    filter(sampleType == "carabid" | sampleType == "other carabid") %>% 
    select(domainID, siteID, plotID, setDate, collectDate, 
           sampleID, subsampleID, taxonRank, taxonID, scientificName, 
           morphospeciesID, individualCount, remarks) %>%
    left_join(bet_fielddata %>%
                  select(domainID, siteID, plotID, trapID, setDate, 
                         collectDate, sampleCollected, sampleID)) %>% #use bet_fielddata's trapID since there are no NA's
    mutate(trapnights = as.numeric(collectDate - setDate, units="days"),
           col_DOY = yday(collectDate),
           col_month = month(collectDate),
           col_year = year(collectDate),
           spec_morph = ifelse(is.na(morphospeciesID), #merge parataxonomist species and morphospecies columns
                               scientificName, 
                               paste0(scientificName,"/", morphospeciesID,sep="")))

saveRDS(carabid_clean, "data/carabid_clean.rds")
```

### Outstanding data questions
* Is collectionDate a reliable index for collection bouts?
