---
title: "abundance"
author: "Anna Spiers"
date: "7/9/2020"
output: html_document
---

A workflow for creating dummy forecasts for NEON carabid abundance.  
1) EDA   
2) Average baseline dummy forecast  
3) Random walk dummy forecast 

```{r, message=F}
library(neonUtilities)
library(ggplot2)
library(dplyr)
library(stringr) #word()
library(lubridate)
```

#### Load data

```{r}
# dataframe with comprehensive information - more than is needed for forecasting abundance
carabid_clean <- readRDS("data/carabid_clean.rds") 
# Each row has a unique subsampleID. the individualCount column reports the raw count 
# of individuals for a subsample (taxon per trap per collectDate)

# dataframe with information related to forecasting abundance only
carabid_count <- carabid_clean %>%
    select(siteID, sampleID, collectDate, col_DOY, col_month, col_year, individualCount, trapnights) %>%
    group_by(siteID, collectDate, trapnights) %>%
    summarize(totalCount = sum(individualCount))
```

#### EDA

See [Carl's script](https://github.com/boettiger-lab/neon-forecast/blob/master/abundance/beetles.Rmd) exploring abundance, particularly raw count vs abundance, for more information.

Count at each site across collection bouts
```{r}
bet_sorting %>%
    filter(sampleType == "carabid" | sampleType == "other carabid") %>%
    select(domainID, siteID, plotID, trapID, setDate, collectDate, 
           individualCount, sampleID, subsampleID) %>%
    mutate(plot_trap = paste0(plotID, "_", trapID, sep=""),
           trapnight = as.numeric(collectDate - setDate, units="days"),
           col_DOY = yday(collectDate),
           year = as.character(year(collectDate))) %>%
    group_by(siteID, year, col_DOY) %>%
    summarize(n=sum(individualCount)) %>%
    ggplot() + 
        geom_line(aes(x=col_DOY, y=n, col=year)) +
    facet_wrap(~siteID)
```

Count at each site across years
```{r}
bet_sorting %>%
    filter(sampleType == "carabid" | sampleType == "other carabid") %>%
    select(domainID, siteID, plotID, trapID, setDate, collectDate, 
           individualCount, sampleID, subsampleID) %>%
    mutate(plot_trap = paste0(plotID, "_", trapID, sep=""),
           trapnight = as.numeric(collectDate - setDate, units="days"),
           year = year(collectDate)) %>%
    group_by(siteID, year) %>%
    summarize(n=sum(individualCount)) %>%
    ggplot(aes(x=year, y=n)) + 
        geom_line() +
    facet_wrap(~siteID)
```










####Average baseline dummy forecast  

```{r}

mnCarabid <- carabid_count %>%
	group_by(siteID) %>% 
	summarise(mnAbund=mean(totalCount, na.rm=TRUE), 
		mnTrapNight=mean(trapnights, na.rm=TRUE), 
		mnSAbund=mean(totalCount/trapnights, na.rm=TRUE), 
		sdAbund=sd(totalCount, na.rm=TRUE),
		sdSabund=sd(totalCount/trapnights, na.rm=TRUE))

```








#### Random walk dummy forecast 

```{r}

rwCarabid <- carabid_count %>%
	group_by(siteID) %>% 
	summarise(rw.mnAbund=mean(totalCount[which.max(collectDate)], na.rm=TRUE), 
		rw.TrapNight=mean(trapnights[which.max(collectDate)], na.rm=TRUE), 
		rw.SAbund=mean(totalCount[which.max(collectDate)]/
			trapnights[which.max(collectDate)], na.rm=TRUE), 
		rw.sdAbund=sd(totalCount[which.max(collectDate)]),
		rw.sdSabund=sd(totalCount[which.max(collectDate)]/
			trapnights[which.max(collectDate)], na.rm=TRUE)
	) 


```






