---
output: github_document
---


Helper functions to construct EML, primary to automate the building of attributeList from the data.frame.  

```{r}
source("R/meta.R")
```

Specify core metadata as list objects:

```{r}
carl <- "http://orcid.org/0000-0002-1642-628X"
authors <- list(list(individualName = list(givenName = "Carl", surName = "Boettiger"), 
                id = carl),
                list(individualName = list(givenName = "Anna",  surName ="Spiers")),
                list(individualName = list(givenName = "Tad", surName = "Dallas")),
                list(individualName = list(givenName = "Brett", surName = "Melbourne")),
                list(individualName = list(givenName = "Kari", surName = "Norman")),
                list(individualName = list(givenName = "Eric", surName = "Sokol")),
                list(individualName = list(givenName = "Kelsey", surName = "Yule")),
                list(individualName = list(givenName = "Nico", surName = "Franz")),
                list(individualName = list(givenName = "Thilina", surName = "Surasinghe")),
                list(individualName = list(givenName = "Quinn", surName = "Thomas"))
                )

tables <- list(
  list(file = "beetle-targets.csv.gz", 
       description = "Measured carabid beetle species richness and abundance
          by siteID and collectDate across all NEON sites operating pitfall traps"),
  list(file = "beetle-forecast-null_average.csv.gz", 
       description = "Forecast of carabid beetle species richness and abundance
          by siteID and collectDate across all NEON sites operating pitfall traps")
)
```


Pre-compute geographic coverage from NEON EML, and also generate taxonomic coverage.  Parsing thousands of NEON EML is inefficient, we really need only one EML per site, so would be better if we restricted the file list to a YYYY-MM when all sites were sampling. To streamline things, we store the coverage data instead.  (Alternately we could have constructed the geographicCoverage from the NEON API locations endpoint).  Note that NEON EML lists each NEON site as effectively a point location, despite EML being defined as a bounding box.  

```{r eval=FALSE}
library(dplyr)
library(jsonlite)
meta <- neonstore::neon_index(ext="xml", product = "DP1.10022.001")
all <- lapply(meta$path, emld::as_emld)
geo <- lapply(all, function(x) x$dataset$coverage$geographicCoverage)
geo %>% toJSON() %>% fromJSON() %>% distinct() %>% write_json("meta/bet_geo.json", auto_unbox=TRUE)
```

```{r eval = FALSE}
library(dplyr)
library(jsonlite)
library(EML)
taxonomicCoverage <- EML::set_taxonomicCoverage(
  data.frame(Kingdom = "Animalia", Phylum="Insecta", Class = "Coleoptera", Family="Carabidae"))
taxonomicCoverage %>% jsonlite::write_json("meta/carabid_taxa.json")
```

Load the stored taxonomic and geographic coverage, compute current temporal coverage:

```{r}
target <- vroom::vroom("beetle-targets.csv.gz")
dates <- as.Date(paste(target$year, target$month, "01", sep="-"), format = "%Y-%b-%d")
startDate <- min(dates)
endDate <- max(dates) # Or is this the forecast horizon
temporalCoverage <- 
              list(rangeOfDates =
                list(beginDate = list(calendarDate = startDate),
                     endDate = list(calendarDate = endDate)))
                     
geographicCoverage <- jsonlite::read_json("meta/bet_geo.json")
taxonomicCoverage <-  jsonlite::read_json("meta/carabid_taxa.json")

coverage <- list(geographicCoverage = geographicCoverage,
                 temporalCoverage = temporalCoverage,
                 taxonomicCoverage = taxonomicCoverage)
               
```


Now generate EML. Note that `attributesList` will be computed from the `vroom::spec` of the data files,
which handles data types reasonably but is fast and loose on units.  Fortunately `dimensionless` is 
okay for the count data included here. 

```{r}
meta <- build_eml(title = "NEON Carabid Species Richness and Abundance Forecast", 
          abstract = "Simple forecast of Carabid beetle species richness and abundance at
                     each month at each NEON site for 2019, based on historical averages.", 
          creators = authors, 
          contact_orcid = carl,
          coverage = coverage,
          tables = tables)


emld::as_xml(meta, "meta/eml.xml")
emld::eml_validate("meta/eml.xml")
```


## Publish the product to DataONE:

Helper functions which wrap around the standard `dataone` R package functions.
Main difference is to default to content-based identifiers (hash URIs) for each
object, and to set sha256 as the checksum used for objects.  Also adds relationships
and provenance.  


To actually upload to the production KNB server, you will need to have a `dataone_token` defined in `options()`. For this to run in test mode, you will need to have a `dataone_test_token` defined in `options()`.  Having a test token available forces use of the dataone testing server, even if a production token is also available.  

```{r}
source("R/dataone.R")
```

```{r eval = FALSE}
publish_dataone(in_file="products/richness.csv", 
                out_file="products/richness_forecast.csv", 
                code="workflow.Rmd", 
                meta="meta/eml.xml",
                orcid=carl)
```


