---
title: "abundance"
author: "Anna Spiers"
date: "7/9/2020"
output: html_document
---

A workflow for creating dummy forecasts for NEON carabid species richness  
1) EDA
2) Average baseline dummy forecast  
3) Random walk dummy forecast 

```{r, message=F}
library(neonUtilities)
library(ggplot2)
library(plyr)
library(dplyr)
library(stringr) #word()
library(lubridate)
```

#### Load data

```{r}
# dataframe with comprehensive information - more than is needed for forecasting richness
carabid_clean <- readRDS("data/carabid_clean.rds") 

# dataframe with information related to forecasting richness only
carabid_rich <- carabid_clean %>%
    select(siteID, sampleID, collectDate, col_DOY, col_month, col_year, taxonRank,
           taxonID, scientificName, morphospeciesID, spec_morph) %>%
    count(siteID, collectDate) 
# n is the number of unique scientific names and morphospecies at a site on a collectiondate
```

#### EDA

Where are morphospecies assigned? I thought that morphospecies were assigned when the parataxonomist couldn't ID to species, but this may be wrong. May be worth looking at bet_parataxonomist data again, as my hypothesis was the case there, at least.




# Visualize spatial and temporal accumulation curves, and Chao's index



```{r}

sum(is.na(filter(carabid_clean, taxonRank=='family')$morphospeciesID))

sum(!is.na(filter(carabid_clean, taxonRank=='family')$morphospeciesID))

```



Plot of time series of richness as function of collect date and siteID

```{r}

pdf('richnessOverTime.pdf')
par(mar=c(4,4,0.5,0.5))
with(dplyr::filter(carabid_rich, siteID=='ABBY'), 
	plot(collectDate, n, ylim=c(0, 215), 
		xlim=range(carabid_rich$collectDate),
		xlab='Collection date', ylab='Richness', 
		type='n')
)

sites <- unique(carabid_rich$siteID)

for(i in 1:length(sites)){
	tmp <- dplyr::filter(carabid_rich, siteID==sites[i])
	tmp <- tmp[order(tmp$collectDate),]
	lines(tmp$collectDate, tmp$n, col=rainbow(50)[i], lwd=1.25)
}
dev.off()

```




### Site by species matrices


Currently visualized for every year, as I could not find a unique sampling identifier corresponding to site (collectDate does not work, maybe some combination of col_month and siteID?)

The values in the cells are also not abundances. They are the number of times that species was observed during the year in a given site. 


```{r}

yearz <- unique(carabid_clean$col_year)

intMats <- list()

for(i in 1:length(yearz)){
	tmp <- dplyr::filter(carabid_clean, col_year==yearz[i])
	tmp <- as.matrix(table(tmp$siteID, tmp$scientificName))
	tmp[tmp==0] <- NA
	intMats[[i]] <- tmp
}

```




```{r}

pdf('intMats.pdf')
layout(matrix(1:8, ncol=2,nrow=4))
par(mar=c(1,1,1,1))
lapply(intMats, image, col=viridis::viridis(100))
dev.off()

```


Considering the mean abundance instead of the number of times a species was recorded. 

```{r}

yearz <- unique(carabid_clean$col_year)

intMats2 <- list()


for(i in 1:length(yearz)){
	tmp <- dplyr::filter(carabid_clean, col_year==yearz[i])
	tmp <- tmp %>%
		group_by(siteID, scientificName) %>%
		summarise(mnNum = mean(individualCount, na.rm=TRUE), 
			sumNum = sum(individualCount, na.rm=TRUE), 
			sdNum = sd(individualCount, na.rm=TRUE))
	intMats2[[i]] <- reshape2::acast(tmp, siteID ~ scientificName, 
		value.var='mnNum')
}


```




```{r}
pdf('intMats2.pdf')
layout(matrix(1:8, ncol=2,nrow=4))
par(mar=c(1,1,1,1))
lapply(intMats, image, col=viridis::viridis(100))
dev.off()
```






### How much does a single plot matter?

Focus on a single sampling bout and a single site

```{r}

destroyPlot <- function(dat, numPlots=2){
	plotz <- unique(dat$plotID)
	keep <- sample(plotz, length(plotz) - numPlots)
	dat2 <- dplyr::filter(dat, plotID %in% keep)
	return(c(rich=length(unique(dat$scientificName)), 
		poor= length(unique(dat2$scientificName))))
}


# pick a random site
dat <- dplyr::filter(carabid_clean, siteID==sample(carabid_clean$siteID, 1))
# pick a random setDate
dat <- dplyr::filter(dat, setDate==sample(dat$setDate, 1))

plyr::ldply(1:5, function(x){destroyPlot(dat, numPlots=x)})

```

Run it a few times, to check how much excluding an entire plot influences overall richness estimates. It does, but not nearly to the extent that one would think. 

















####Average baseline dummy forecast  


```{r}

mnCarabid <- carabid_rich %>%
	group_by(siteID) %>% 
	summarise(mnRichness=mean(n, na.rm=TRUE), 
		sdRichness=sd(n, na.rm=TRUE)
)

```








#### Random walk dummy forecast 

I do not think the error is right here?

```{r}

rwCarabid <- carabid_rich %>%
	group_by(siteID) %>% 
	summarise(rw.mnRichness=n[which.max(collectDate)], 
		rw.sdRichness=sd(n, na.rm=TRUE)
	)

```















