---
title: "abundance"
author: "Anna Spiers"
date: "7/9/2020"
output: html_document
---

A workflow for creating dummy forecasts for NEON carabid species richness  
1) EDA
2) Average baseline dummy forecast  
3) Random walk dummy forecast 

```{r, message=F}
library(neonUtilities)
library(ggplot2)
library(dplyr)
library(stringr) #word()
library(lubridate)
```

#### Load data

```{r}
# dataframe with comprehensive information - more than is needed for forecasting richness
carabid_clean <- readRDS("data/carabid_clean.rds") 

# dataframe with information related to forecasting richness only
carabid_rich <- carabid_clean %>%
    select(siteID, sampleID, collectDate, col_DOY, col_month, col_year, taxonRank,
           taxonID, scientificName, morphospeciesID, spec_morph) %>%
    count(siteID, collectDate) 
# n is the number of unique scientific names and morphospecies at a site on a collectiondate
```

#### EDA

Where are morphospecies assigned? I thought that morphospecies were assigned when the parataxonomist couldn't ID to species, but this may be wrong. May be worth looking at bet_parataxonomist data again, as my hypothesis was the case there, at least.




# Visualize spatial and temporal accumulation curves, and Chao's index



Plot of time series of richness as function of collect date and siteID

```{r}

pdf('richnessOverTime.pdf')
par(mar=c(4,4,0.5,0.5))
with(dplyr::filter(carabid_rich, siteID=='ABBY'), 
	plot(collectDate, n, ylim=c(0, 215), 
		xlim=range(carabid_rich$collectDate),
		xlab='Collection date', ylab='Richness', 
		type='n')
)

sites <- unique(carabid_rich$siteID)

for(i in 1:length(sites)){
	tmp <- dplyr::filter(carabid_rich, siteID==sites[i])
	tmp <- tmp[order(tmp$collectDate),]
	lines(tmp$collectDate, tmp$n, col=rainbow(50)[i], lwd=1.25)
}
dev.off()

```


















####Average baseline dummy forecast  

```{r}

mnCarabid <- carabid_rich %>%
	group_by(siteID) %>% 
	summarise(mnRichness=mean(n))

```








#### Random walk dummy forecast 

```{r}

rwCarabid <- carabid_rich %>%
	group_by(siteID) %>% 
	summarise(rwRichness=n[which.max(collectDate)])

```















