---
title: "abundance"
author: "Anna Spiers"
date: "7/9/2020"
output: html_document
---

A workflow for creating dummy forecasts for NEON carabid species richness  
1) EDA
2) Average baseline dummy forecast  
3) Random walk dummy forecast 

```{r, message=F}
library(neonUtilities)
library(ggplot2)
library(dplyr)
library(stringr) #word()
library(lubridate)
```

#### Load data

```{r}
# dataframe with comprehensive information - more than is needed for forecasting richness
carabid_clean <- readRDS("data/carabid_clean.rds") 

# dataframe with information related to forecasting richness only
carabid_rich <- carabid_clean %>%
    select(siteID, sampleID, collectDate, col_DOY, col_month, col_year, taxonRank,
           taxonID, scientificName, morphospeciesID, spec_morph) %>%
    count(siteID, collectDate) 
# n is the number of unique scientific names and morphospecies at a site on a collectiondate
```

#### EDA

Where are morphospecies assigned? I thought that morphospecies were assigned when the parataxonomist couldn't ID to species, but this may be wrong. May be worth looking at bet_parataxonomist data again, as my hypothesis was the case there, at least.

# Visualize spatial and temporal accumulation curves, and Chao's index


####Average baseline dummy forecast  



#### Random walk dummy forecast 




