---
title: "scoring.Rmd"
output: github_document
---

```{r setup, message=FALSE}
library(contentid)
library(tidyverse)
```

Load the forecast and true values as determined by `workflow.md`:

```{r}
richness_forecast <- read_csv(resolve("hash://sha256/93e741a4ff044319b3288d71c71d4e95a76039bc3656e252621d3ad49ccc8200"))
true_richness <- read_csv(resolve("hash://sha256/b363cc598b55b4645941c99958542aa16db77363ded84544eb49c6ffe478441e"))
```

Create MCMC replicates from the moments in the forecast:

```{r}


ids <- richness_forecast %>% mutate(id = paste(siteID, year, month, sep="-")) %>% pull(id)

richness_reps <- map_dfr(seq_along(ids), function(i) 
  data.frame(id = ids[i],
             rep = 1:500, 
             y = rnorm(500, richness_forecast$mean[[i]], richness_forecast$sd[[i]])))
```




```{r}
## assumes true_df has column 'id' to join, and 'true' with observed value
## predicted_df has replicates, grouped by 'id' indicating a unique prediction 
## (e.g. n reps at given site and time), with value labeled as `y`
crps_score <- function(predicted_df,
                       true_df
                      ){
   scoring_fn <- function(y, dat) tryCatch(scoringRules::crps_sample(y, dat), error = function(e) NA_real_, finally = NA_real_)
   left_join(predicted_df, true_df)  %>% 
    group_by(id) %>% 
    summarise(score = scoring_fn(y = true[[1]], dat = y))
}

```



```{r}

obs <- true_richness %>% mutate(id = paste(siteID, 2019, month, sep="-")) %>% group_by(id) %>% summarize(true = mean(true))

scores_crps <- crps_score(richness_reps, obs)
```



## Using moment closure, Gneiting & Raferty Eq 27:


```{r}
## note, bigger scores are better here
score <- function(predicted_df,
                  true_df,
                  scoring_fn =  function(x, mu, sigma){ -(mu - x )^2 / sigma^2  - log(sigma)}
                  ){
  true_df %>% 
  left_join(predicted_df)  %>%
  mutate(score = scoring_fn(true, mean, sd))
}

```  


```{r message=FALSE}
richness_score <- score(richness_forecast, true_richness)

## average the scores when we have 2 observations in the same month(?)
eq27 <- richness_score  %>% mutate(id = paste(siteID, year, month, sep="-")) %>%
  group_by(id) %>% summarise(score = mean(score))

```

## Compare best & worst predictions under the two scoring metrics:

Combined scores, sorting by best CRPS score (using `neg_crps` so that bigger is better in both cases)

```{r}
combined <- scores_crps %>% 
  mutate(neg_crps = - score) %>% 
  select(-score) %>%
  left_join(eq27, by="id") %>%
  rename(eq27 = score) %>%
  arrange(desc(neg_crps))

combined
```



Best-predicted observations:

```{r}
# CRPS, small is best
scores_crps %>% arrange(score)
# eq27: Bigger is better:
eq27 %>% arrange(desc(score))
```

Worst-predicted:

```{r}
scores_crps %>% arrange(desc(score))
eq27 %>% arrange(score)
```





